---
title: "Lecture 1: Introduction, Kaplan–Meier and RMST"
toc: true
toc-depth: 3
---

```{r}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)
```



## Learning objectives

::: callout-note
By the end of this lecture, you should be able to:

-   Describe what makes survival data different from standard outcomes
-   Explain how Kaplan–Meier survival curves are constructed
-   Interpret censoring in survival analysis
-   Identify the limitations of Kaplan–Meier curves
-   Understand the motivation for Restricted Mean Survival Time (RMST)
-   Interpret RMST differences and ratios in applied settings
:::

## What is survival data?

Survival data measure the **time until a specific event occurs**.

Examples in kidney research include:\
- Time from transplantation to graft failure\
- Time from transplant to patient death\
- Time to acute rejection

::: callout-note
### Key feature of survival data

The outcome is **composite**: - a **time variable** - an **event indicator**\
(1 = event occurred, 0 = censored)

These two components must always be analysed **together**.
:::

## Kaplan–Meier survival curves

A Kaplan–Meier (KM) curve estimates the **probability of remaining event-free beyond a given time**.

The **KM method** is a **non-parametric** approach used to estimate the
survival function.

- It estimates the probability of surviving beyond a given time.
- It appropriately handles **censoring**, where some individuals have not experienced
  the event by the end of follow-up.

The Kaplan–Meier estimator at time $t_i$ is defined as:

$$
\hat{S}(t_i) = \hat{S}(t_{i-1}) \times \left( 1 - \frac{d_i}{r_i} \right)
$$

where:

- $ d_i $ is the **number of events** occurring at time $ t_i $
- $ r_i $ is the **number of individuals at risk** just before time $ t_i $




::: callout-tip
### How to read a Kaplan–Meier curve

-   The curve decreases in **steps**, not smoothly
-   Each drop corresponds to an **event time**
-   Flat sections indicate **no events**
-   Censored observations:
    -   do **not** cause the curve to drop
    -   reduce the number at risk going forward
:::

## Example: simple Kaplan–Meier estimate
This dataset captures the time from kidney transplantation to graft failure. The only predictor variable included is donor type.

| ID | Graft failure | Time to graft failure (months) | Donor type |
|----|---------------|--------------------------------|------------|
| 1  | 1             | 3                              | Living     |
| 2  | 0             | 5                              | Deceased   |
| 3  | 1             | 6                              | Living     |
| 4  | 1             | 8                              | Deceased   |
| 5  | 0             | 10                             | Living     |
| 6  | 1             | 12                             | Deceased   |



The KM estimates can be written as follows:  

| Sorted time | Graft failure | Number of people at risk | Survival probability |
|------------|---------------|--------------------------|----------------------|
| 0          | –             | 6                        | 1.000                |
| 3          | 1             | 6                        | 1 × (1 − 1/6) = 0.833 |
| 6          | 1             | 4                        | 0.833 × (1 − 1/4) = 0.625 |
| 8          | 1             | 3                        | 0.625 × (1 − 1/3) = 0.417 |
| 12         | 1             | 2                        | 0.417 × (1 − 1/1) = 0 |


```{r,warning=FALSE}

library(survival)
library(survminer)

df <- data.frame(
  id = 1:6,
  time = c(3, 5, 6, 8, 10, 12),
  status = c(1, 0, 1, 1, 0, 1)
)

km_fit <- survfit(Surv(time, status) ~ 1, data = df)

ggsurvplot(
  km_fit,
  data = df,
  censor = TRUE,
  risk.table = TRUE,
  xlab = "Time from transplant (months)",
  ylab = "Survival probability",
  ggtheme = theme_minimal()
)
```

::: {.callout-tip}
### Interpretation

At any given time point, the survival probability represents the estimated proportion
of individuals who remain event-free beyond that time.

Censored individuals:  
- contribute information up to their censoring time  
- do not count as failures  
- do not cause drops in the curve  
:::


## Kaplan–Meier curves by donor type

The simulated data contains the time from kidney transplantation to graft failure. The only predictor variable included is donor type.


```{r,warning=FALSE}
set.seed(2025)

n <- 200
donor_type <- sample(c("Living", "Deceased"), n, replace = TRUE)

time <- ifelse(donor_type == "Living",
               rexp(n, rate = 0.05),
               rexp(n, rate = 0.1))

censoring_time <- rexp(n, rate = 0.05)
observed_time <- pmin(time, censoring_time)
status <- as.integer(time <= censoring_time)

df <- data.frame(
  time = round(observed_time, 1),
  status = status,
  donor_type = donor_type
)

km_fit <- survfit(Surv(time, status) ~ donor_type, data = df)

ggsurvplot(
  km_fit,
  data = df,
  risk.table = TRUE,
  pval = TRUE,
  conf.int = TRUE,
  legend.title = "Donor type",
  xlab = "Time to graft failure (months)",
  ylab = "Survival probability",
  ggtheme = theme_minimal()
)
```


## Limitation of Kaplan–Meier curves

::: {.callout-warning}
Kaplan–Meier curves allow **group comparison only**.

They cannot:  
- adjust for confounders  
- include continuous predictors  

For adjusted analyses, we need regression-based approaches,
such as Cox regression or **RMST regression**.
:::

## Restricted Mean Survival Time (RMST)

So far, we’ve seen that **Kaplan-Meier** does not adjust for confounders.
To handle covariates, we need approaches like:  
- Cox regression  
- RMST regression  
- Parametric survival models  

In censored data, the true mean survival time is often not observable. We don’t have event time for some of the individual they may not experience the event.

Because of censoring, estimating the unrestricted mean survival time is often not feasible.
Instead, we define a fixed time point $𝝉$ and focus on Restricted Mean Survival Time (RMST):  
The average time survived up to time $𝝉$. 



::: {.callout-note}
RMST is the **average time survived up to a fixed time point τ**.

It answers the question:

> “How long, on average, do individuals remain event-free within τ months?”
:::

::: {.callout-tip}
### Why RMST is attractive

- Directly interpretable
- Does not require proportional hazards
- Works well when survival curves cross
:::




```{r, message=FALSE, warning=FALSE}
library(ggplot2)

# Parameters
tau <- 5
rate <- 0.2
tmax <- 10


# Example survival curve
# S(t) = exp(-rate * t)

surv_df <- data.frame(time = seq(0, tmax, by = 0.01))
surv_df$S <- exp(-rate * surv_df$time)

# Area under curve up to tau (RMST region)
surv_df_shade <- subset(surv_df, time <= tau)


# Plot
ggplot(surv_df, aes(x = time, y = S)) +
  geom_area(
    data = surv_df_shade,
    aes(x = time, y = S),
    fill = "skyblue",
    alpha = 0.35
  ) +
  geom_line(
    linewidth = 1.2,
    colour = "blue"
  ) +
  geom_vline(
    xintercept = tau,
    linetype = "dashed",
    colour = "red",
    linewidth = 0.9
  ) +
  annotate(
    "text",
    x = tau + 0.15,
    y = 0.5,
    label = paste0("Tau = ", tau),
    colour = "red",
    size = 5,
    hjust = 0
  ) +
  labs(
    x = "Time",
    y = "Survival Probability"
  ) +
  coord_cartesian(xlim = c(0, tmax), ylim = c(0, 1)) +
  theme_minimal(base_size = 14)
```





## Adjusted RMST regression in Kidney Transplant Patients


We aim to estimate the **restricted mean survival time (RMST)** up to **$\tau = 36$ months** after kidney transplantation, and to compare outcomes between donor types while adjusting for important clinical factors.  


Outcome  
- **Time to graft failure** (months since transplantation)  

Event Indicator  
- **Graft failure status**
  - 1 = graft failure occurred  
  - 0 = censored (no failure by end of follow-up)  

Main Exposure  
- **Donor type**
  - Living donor  
  - Deceased donor    


Covariates (Adjusted For)  
To reduce confounding, we adjust for:  
- **Donor age** (years)  
- **Recipient gender** (male / female)  
- **Baseline eGFR at transplant** (kidney function at time of surgery)  

Interpretation Goal  
After adjustment, the RMST contrast represents:  

> The **average difference in graft-survival time up to $36$ months** between living and deceased donor transplants,  
> among patients with similar donor age, recipient gender, and baseline eGFR.




```{r,warning=FALSE, message=FALSE}


library(survival)
library(survRM2)
library(survminer)
library(dplyr)


# -------------------- Simulate the dataset

set.seed(2025)
n <- 1000

# Donor type
donor_type <- sample(c("Living", "Deceased"), size = n, replace = TRUE)

# Donor age (older for deceased donors)
donor_age <- ifelse(donor_type == "Living",
                    rnorm(n, mean = 45, sd = 10),
                    rnorm(n, mean = 60, sd = 10))
donor_age <- round(pmax(donor_age, 18))

# Recipient sex
recipient_sex <- sample(c("Male", "Female"), size = n, replace = TRUE)

# eGFR at transplant
egfr_at_tx <- round(runif(n, min = 5, max = 60), 1)

# Simulated survival times (longer for living donors)
time <- ifelse(donor_type == "Living",
               rexp(n, rate = 0.05),  # mean ~20
               rexp(n, rate = 0.1))   # mean ~10

# Censoring
censoring_time <- rexp(n, rate = 0.05)
#observed_time <- pmin(time, censoring_time)
observed_time <- pmax(pmin(time, censoring_time), 0.01)
status <- as.integer(time <= censoring_time)

# Create dataset
df <- data.frame(
  id = 1:n,
  time = round(observed_time, 1),
  status = status,
  donor_type = donor_type,
  donor_age = donor_age,
  recipient_sex = recipient_sex,
  egfr_at_tx = egfr_at_tx
)

# Encode donor_type as binary: 1 = Living, 0 = Deceased
df$donor_type_bin <- ifelse(df$donor_type == "Living", 1, 0)


#----------------- Fit RMST Regression (with covariates)

# Set truncation time τ (e.g., 36 months)
tau <- 36

df$recipient_sex <- as.factor(df$recipient_sex)
df$donor_type_bin <- ifelse(df$donor_type == "Living", 1, 0)
df$time[df$time == 0] <- 0.01
covariates <- model.matrix(~ donor_age + recipient_sex + egfr_at_tx, data = df)[, -1]
str(df$recipient_sex)

# Fit RMST regression model
rmst_model <- rmst2(time = df$time,
                    status = df$status,
                    arm = df$donor_type_bin,
                    tau = 36,
                    covariates = covariates
                    )

rmst_model



```

::: {.callout-tip}
### Interpretation (RMST difference)

RMST (Living donor - Deceased donor) = 6.30 months95% CI: 3.91 to 8.68, p < 0.001
Interpretation: On average, recipients of living donor kidneys survive 6.3 more months event-free (up to 36 months) than those with deceased donors, after adjusting for age, sex, and eGFR at transplant.

:::

::: {.callout-tip}
### Interpretation (RMST ratio)

RMST ratio = 1.69 (Living vs Deceased) 95% CI: 1.39 to 2.05, p < 0.001
Over the 36-month period, recipients of living donor kidneys have 69% longer average event-free survival time than recipients of deceased donor kidneys, after adjustment.

:::

::: {.callout-tip}
### Interpretation (Covariate coefficients)

None of the covariates were significant in our simulated data but if for an example egfr was significant then the interpretation would be: 

Each unit increase in eGFR at the time of transplant is associated with an additional 0.047(coefficient in the model ) months (≈3 days) of average event-free survival time over 36 months, adjusting for donor type, donor age, and recipient sex.

:::

::: {.callout-question}

**Think about the following before moving on:**


- Why do we need to choose a truncation time $τ$?
- What happens if τ is too short or too long?
- When would RMST be preferable to a Cox model?
:::












