---
title: "Lecture 3: Cox model when PH assumption is violated"
toc: true
toc-depth: 3
---


```{r, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE
)
```

## Learning Objectives

::: callout-note
By the end of this lecture, you should be able to:

- Recognise when the **proportional hazards (PH) assumption** is violated
- Understand different strategies to handle non-PH:
  - Stratified Cox model
  - Time-varying effects
  - Extended Cox with time-varying covariates
- Understand when a **joint model** is needed

:::



## Reminder: Proportional Hazards Assumption

The Cox model assumes that **hazard ratios are constant over time**.

This means that if one group has higher risk than another at the beginning,
it has the same *relative* risk later in follow-up.

::: {.callout-note}
If this assumption does not hold, the estimated hazard ratio from a standard Cox model
may be misleading.
:::


## What If Proportional Hazards Does Not Hold?

There are several approaches when PH is violated:

1. **Stratified Cox model:** allow baseline hazards to differ by group.
2. **Include time-varying effect:** model changing effects over time, include interaction between covariate and time (e.g. $𝑥×𝑡$). 
3. **Use extended Cox model:** allow a covariate’s value itself to change over time.



## Stratified Cox model

When the proportional hazards (PH) assumption is violated for a categorical variable, one option is a **stratified Cox model**.

A stratified Cox model can be written as:

$$
h_s(t \mid X) = h_{0s}(t)\,\exp(\beta_1 X_1 + \beta_2 X_2 + \cdots)
$$

where:

- $s$ is the stratum (e.g. donor type)
- $h_{0s}(t)$ is the **baseline hazard for stratum $s$**
- $X_1, X_2, \ldots$ are other covariates

 **What does stratification mean?**  

If we stratify on a variable (e.g. **donor type**):

- We **do not estimate a single hazard ratio** for the stratifying variable anymore  
  (we are not saying “living vs deceased = HR 0.7”)
- The stratifying variable is used only to **define strata**
- Each stratum is allowed to have a **different baseline risk trajectory** over time

In words:

> Patients with living donors and patients with deceased donors are allowed to have completely different baseline hazards over time.

**What is still estimated?**   

We still estimate the effects of other covariates (e.g. donor age, recipient sex, eGFR at transplant), assuming those effects are **proportional within each stratum**.

::: {.callout-note}
Stratification is useful when a variable strongly affects baseline risk but does not satisfy the PH assumption.
:::


Stratification can be thought of as:

> **“Parallel Cox models with different baseline hazards, but the same slopes for the other covariates.”**

That is, each group has its own baseline risk over time, but covariate effects are assumed to be the same across groups.

---

 ✅  **When to use stratification**
Use stratification when:

- The proportional hazards (PH) violation is mainly driven by **one categorical variable** (e.g. donor type)
- You still want to **adjust for this variable**
- But you do **not need its hazard ratio** as a main result

In this case, the variable is used to define strata rather than being estimated as a coefficient.

---

❌ **When not to use stratification**
Do **not** use stratification when:

- The stratifying variable is your **primary exposure of interest**
- You need to report a **hazard ratio comparing groups**

Because once you stratify on a variable (e.g. donor type), you **no longer obtain a single hazard ratio** comparing those groups.

## Example: Handling Non-Proportional Hazards Using Stratified Cox Model

We are interested in whether time to graft failure after kidney transplantation differs by donor type (living vs deceased), after adjusting for important clinical factors.


We begin by fitting a standard Cox proportional hazards model, then check the PH assumption.
If the PH assumption is violated for donor type, we refit the model using stratification by donor type.

```{r, message=FALSE, warning=FALSE}
library(survival)

stratified_kidney_data<-read.csv("stratified_kidney_data.CSV")

cox_standard <- coxph(
  Surv(time, status) ~ donor_type +
    donor_age +
    recipient_gender +
    egfr_baseline,
  data = stratified_kidney_data
)

#summary(cox_standard)

#check the PH assumption first 
ph_test <- cox.zph(cox_standard)
ph_test
plot(ph_test)
```

::: {.callout-note}
A small p-value indicates evidence against the proportional hazards assumption.
:::

**Interpretation of PH diagnostics**  

The proportional hazards assumption is violated for donor type.  
Other covariates do not show strong evidence of violation.  
The global test is significant, indicating overall non-proportionality.  
This suggests that the effect of donor type on graft failure risk changes over time, making the standard Cox model inappropriate.  


**Stratified Cox model**  

To address this violation, we fit a stratified Cox model, stratifying on donor type.

```{r, message=FALSE, warning=FALSE}
library(survival)
cox_strat <- coxph(
Surv(time, status) ~ donor_age + recipient_gender + egfr_baseline +
strata(donor_type),
data = stratified_kidney_data
)

summary(cox_strat)

```


::: {.callout-note}
- We no longer estimate a hazard ratio for donor type.

- Donor type is now used only to define separate baseline hazards.

- Effects of other covariates are assumed proportional within each stratum.
:::


🧠  **Interpretation of stratified Cox model**

Within each donor type (living and deceased), we interpret covariate effects as follows:

✅ Donor age  

HR ≈ 1.01 (95% CI: 0.997–1.03, P-value=0.14), Not statistically significant.  
Within each donor type, donor age shows no strong association with graft failure.

✅ Recipient gender (male)

HR ≈ 1.58 (95% CI: 1.16–2.13, P-value=0.003), Statistically significant.
Male recipients have about 58% higher hazard of graft failure compared to females
This comparison is within donor type, adjusting for donor age and eGFR.

✅ eGFR at transplant

HR ≈ 0.99 (95% CI: 0.97–1.00, P-value=0.033), Statistically significant.
Each 1-unit increase in baseline eGFR is associated with about 1–2% lower hazard of graft failure, within each donor type.
To improve interpretability, we can express the effect per **10-unit increase** in eGFR:

$$
HR_{10\text{-unit}} = 0.99^{10} \approx 0.90
$$

**Interpretation:**  
Within each donor type, a 10-unit higher eGFR at the time of transplant is associated with approximately a **10% lower hazard of graft failure**, after adjusting for donor age and recipient gender.


::: {.callout-tip}


### What about donor type?

Stratifying by donor type allows each group to have its own baseline hazard curve, effectively removing the need to assume a constant hazard ratio over time.  

The curves demonstrate that living and deceased donor grafts follow different survival trajectories — exactly what we expect when the PH assumption is violated.

:::

```{r, message=FALSE, warning=FALSE}

# cox_strat is the stratified cox model we fitted in previous chunk 
sf_strat <- survfit(cox_strat)

# Convert survfit to a tidy data frame for ggplot
sf_df <- data.frame(
  time  = sf_strat$time,
  surv  = sf_strat$surv,
  lower = sf_strat$lower,
  upper = sf_strat$upper,
  strata = rep(names(sf_strat$strata), sf_strat$strata)
)

# Clean up the stratum labels from "donor_type=deceased" → "deceased"
sf_df$strata <- sub("donor_type=", "", sf_df$strata)

library(ggplot2)

stratified_cox_plot<-ggplot(sf_df, aes(x = time, y = surv, colour = strata)) +
  geom_step(linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = strata),
              alpha = 0.15, linewidth = 0) +
  scale_colour_manual(values = c("deceased" = "#365abd",  # deep blue
                                 "living"   = "#c0397e"))+ # magenta/rose

  scale_fill_manual(values = c("deceased" = "#365abd",
                               "living"   = "#c0397e")) +
  labs(
    title = "Stratified Cox Model\nDifferent Baseline Hazards by Donor Type",
    x = "Time since transplant (years)",
    y = "Survival probability",
    colour = "Donor type",
    fill   = "Donor type"
  ) +
  coord_cartesian(ylim = c(0.4, 1)) +
  theme_grey(base_size = 14) +
  theme(
    plot.title       = element_text(face = "bold", hjust = 0.5),
    legend.position  = "bottom",
    legend.title     = element_text(face = "bold")
  )

stratified_cox_plot

```




::: {.callout-warning}
### Why Ignoring PH Violations Can Be Misleading

If we ignore the violation of the proportional hazards (PH) assumption and proceed with a standard Cox model, we may obtain a single hazard ratio that does **not accurately reflect how risk changes over time**.

This can lead to misleading conclusions, as the estimated effect represents an average over follow-up rather than the true time-varying relationship.

This is illustrated in the plot below, where the separation between groups is clearly **not constant over time**.

:::


```{r, message=FALSE, warning=FALSE}

#data preparation for plot 

new_deceased <- data.frame(
  donor_type = factor("deceased", levels = levels(as.factor(stratified_kidney_data$donor_type))),
  donor_age = mean(stratified_kidney_data$donor_age, na.rm = TRUE),
  recipient_gender = factor("female", levels = levels(as.factor(stratified_kidney_data$recipient_gender))),
  egfr_baseline = mean(stratified_kidney_data$egfr_baseline, na.rm = TRUE)
)

new_living <- data.frame(
  donor_type = factor("living", levels = levels(as.factor(stratified_kidney_data$donor_type))),
  donor_age = mean(stratified_kidney_data$donor_age, na.rm = TRUE),
  recipient_gender = factor("female", levels = levels(as.factor(stratified_kidney_data$recipient_gender))),
  egfr_baseline = mean(stratified_kidney_data$egfr_baseline, na.rm = TRUE)
)





sf_deceased <- survfit(cox_standard, newdata = new_deceased)
sf_living   <- survfit(cox_standard, newdata = new_living)


df_dec <- data.frame(
  time       = sf_deceased$time,
  surv       = sf_deceased$surv,
  lower      = sf_deceased$lower,
  upper      = sf_deceased$upper,
  donor_type = "Deceased donor"
)

df_liv <- data.frame(
  time       = sf_living$time,
  surv       = sf_living$surv,
  lower      = sf_living$lower,
  upper      = sf_living$upper,
  donor_type = "Living donor"
)


df_ph <- rbind(df_dec, df_liv)




standard_cox_plot<-ggplot(df_ph, aes(x = time, y = surv, colour = donor_type, fill = donor_type)) +
  # CI ribbon
  geom_ribbon(aes(ymin = lower, ymax = upper),
              alpha = 0.2,
              colour = NA,
              show.legend = FALSE) +
  # Step survival curve
  geom_step(linewidth = 1.2) +
  
  # Manual colours for consistency across slides
  scale_colour_manual(
    values = c("Deceased donor" = "#365abd",   # blue
               "Living donor"   = "#c0397e")   # magenta/rose
  ) +
  scale_fill_manual(
    values = c("Deceased donor" = "#365abd",
               "Living donor"   = "#c0397e")
  ) +
  
  # Axis labels and title
  labs(
    title = "Standard Cox PH Model\nAssumes Constant Effect of Donor Type",
    x = "Time since transplant (years)",
    y = "Predicted survival probability",
    colour = "Donor type"
  ) +
  
  # y-axis from 0.4 to 1.0
  coord_cartesian(ylim = c(0.4, 1.0)) +
  
  # Use grey theme with gridlines
  theme_grey(base_size = 14) +
  theme(
    plot.title       = element_text(face = "bold", hjust = 0.5),
    legend.position  = "bottom",
    legend.title     = element_text(face = "bold")
  )




 
library(gridExtra)

grid.arrange(
  standard_cox_plot,
  stratified_cox_plot,
  ncol = 2
)

```


